---
title: "HW3 on Text Wrangling - ANSWER KEY"
author: "Vicki Hertzberg & Melinda Higgins"
date: "02/08/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

### Homework 3 - Text Wrangling

For this homework you will mainly be working with these packages:

* `stringr` - learn more at:
    - [https://stringr.tidyverse.org/](https://stringr.tidyverse.org/)
    - [https://r4ds.had.co.nz/strings.html](https://r4ds.had.co.nz/strings.html)
* `stringi` - learn more at:
    - [https://cran.r-project.org/web/packages/stringi/index.html](https://cran.r-project.org/web/packages/stringi/index.html)
    - [https://r4ds.had.co.nz/strings.html#stringi](https://r4ds.had.co.nz/strings.html#stringi)

Other packages in use:

* `printr` - for making nice tables in output when knitted
* `dplyr` - for BONUS 2 below
* `tidyr` - for BONUS 2 below

```{r setup chunk 0, include=FALSE}
# set knitr options
# show code
knitr::opts_chunk$set(echo = TRUE)
# hide warnings and messages in output
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

# turn on nice tables in output
library(printr)

# load packages
library(stringr)
library(stringi)
```

### Section 1.

Suppose we have a vector `vText` as follows:

```{r Intro 1}
vText <- c('nurse', 'nut', 'ninja', 'nutrient', 'under', 'unusual')
```

We want to write a regular expression that matches `n`, `nu`, `un`, or `unu` in `vText` and replaces the matching patterns with `.`. To do this, the R code is as follows:

```{r}
pattern <- 'u?nu?'
gsub(pattern, '.', vText)
```

#### Problem 1.

Suppose you have another vector vText as follows:

```{r Problem 1}
vText <- c("bubble", "rubble", "trouble", "couple", 
           "bump", "proud")
```

You want to match `bu`, `ub`, `u` and `bub` and replace with `.`.

Write the R code that will make that happen.

Hint: You want to end up with:
  `".ble"   "r.ble"  "tro.le" "co.ple" ".mp"    "pro.d"`

#### Answer to Problem 1.

```{r Answer 1}
#Insert the code to answer the problem here.
pattern <- 'b?ub?'
gsub(pattern, '.', vText)
```

### Section 2.

You have 3 strings of text that you wish to merge. One way to do this is to use the `paste()` function.

```{r Intro 2}
x <- "I AM SAM. I AM SAM. SAM I AM."
y <- "THAT SAM-I-AM! THAT SAM-I-AM! I DO NOT LIKE THAT SAM-I-AM!"
z <- "DO YOU LIKE GREEN EGGS AND HAM?"

paste(x, y, z, collapse = NULL)
```

## BONUS 1

**Extra credit:** What is the difference if you use the `paste0()` function instead of the `paste()` function above?

_INSERT EXTRA CREDIT ANSWER HERE for BONUS 1_

```{r}
x <- "I AM SAM. I AM SAM. SAM I AM."
y <- "THAT SAM-I-AM! THAT SAM-I-AM! I DO NOT LIKE THAT SAM-I-AM!"
z <- "DO YOU LIKE GREEN EGGS AND HAM?"

paste0(x, y, z, collapse = NULL)
```

ANSWER: Using `paste0()` gets rid of spaces between each test string.

#### Problem 2.

Suppose that you now have 4 lines of text as follows:

```{r Problem 2}
W <- "You may write me down in history"
X <- "With your bitter, twisted lies,"
Y <- "You may tread me in the very dirt"
Z <- "But still, like dust, I'll rise."
```

Write the R code below to merge these 3 strings.

#### Answer to Problem 2.

```{r Answer 2}
#Insert the code to answer the problem here.
paste(W, X, Y, Z, collapse = NULL)
```

### Section 3.

An alternative way to merge these text strings is to concatenate them with the `str_c` function from the `stringr` package.

Suppose we want to concatenate the 3 strings we did above, but also NA. We can do this in these two ways. What is the difference? (Answer to yourself)

```{r Intro 3}
paste(X, Y, Z, NA, collapse = NULL)
str_c(X, Y, Z, NA, collapse = NULL)
```

What is the difference between the two results? (Answer to yourself.)

_The function `str_c()` from the `stringr` package handles the `NA` differently. With the `NA` in the list, `str_c()` returns `NA` whereas `paste()` ignores the `NA`._

#### Problem 3.

We now want to concatenate our 4 vectors and NA. Do this using both methods.

```{r Problem 3}
W <- "You may write me down in history"
X <- "With your bitter, twisted lies,"
Y <- "You may tread me in the very dirt"
Z <- "But still, like dust, I'll rise."
```

#### Answer to Problem 3.

```{r Answer 3}
#Insert the code to answer the problem here.

paste(W, X, Y, Z, NA, collapse = NULL)
str_c(W, X, Y, Z, NA, collapse = NULL)
```

### Section 4.

We can use the `str_sub` function to extract parts of strings. Suppose I wanted to extract the last 8 letters of my name.

```{r Intro 4}
myName <- "Vicki Hertzberg"
Length <- str_length(myName)
last7letters <- str_sub(myName, Length-7, Length)
last7letters
```

#### Problem 4. 
Suppose Melinda Higgins wants to extract the last *10* letters of her full name.

```{r Problem 4}
herName <- "Melinda Higgins"
```

Write the code below to extract the last 6 letters of her name.

#### Answer to Problem 4.

```{r Answer 4}
#Insert the code to answer the problem here.
Length <- str_length(herName)
last10letters <- str_sub(herName, Length-9, Length)
last10letters

# or if you did the last 6... sorry for the confusion
Length <- str_length(herName)
last6letters <- str_sub(herName, Length-5, Length)
last6letters
```

### Section 5

Suppose I have a string and I want to split it into unique words based on the occurrence of a separator, as follows:

```{r Intro 5}
myString <- "The_quick_brown_fox_jumped_over_the_lazy_dog"

#the separator is the character "_"

mySeparatedString <- str_split(myString, "_")
mySeparatedString
```

If you look in your environment you will see that `mySeparatedString` is a List of 1.

#### Problem 5.

I want to separate the following string into separate words:

```{r Problem 5}
myNewString <- "Able-Was-I-Ere-I-Saw-Elba"
```

Split this new string into separate words:

#### Answer to Problem 5.

```{r Answer 5}
#Insert the code to answer the problem here.
mySeparatedString <- str_split(myNewString, "-")
mySeparatedString
```

### Section 6.

On another occasion, I need the same string split so that the last word comes off, and the rest remains intact. I can achieve that in the following way:

```{r Intro 6}
myString <- "The_quick_brown_fox_jumped_over_the_lazy_dog"
myNewSplitSpring <- str_split(myString, "_", n=2)
myNewSplitSpring
```

#### Problem 6. 

Suppose we wanted to split off the first "word" from `myNewString`. Again, we have 

```{r Problem 6}
myNewString <- "Now_is_the_time_for_all_good_men_to_come_to_the_aid_of_their_country"
```

Split off the first word but leave the rest intact.

#### Answer 6.

```{r Answer 6}
#Insert the code to answer the problem here.
myNewSplitSpring <- str_split(myNewString, "_", n=2)
myNewSplitSpring
```

### Section 7.

With the `stringi` package there is functionality to count the number of words in a string.

```{r Intro 7}
newString <- "The quick brown fox jumps over the lazy dog."
stri_count_words(newString)
```

#### Problem 7.

```{r}
yourNewString <- "Whether you think you can, or you think you can’t – you’re right,"
```

Use the `stri_count_words` function as above to count the number of distinct words in `yourNewString`.

#### Answer 7.

```{r Answer 7}
#Insert the code to answer the problem here.
stri_count_words(yourNewString)
```

### Section 8.

Let's say you have a string listing famous mathematicians and you want to know if there are any duplicates in the list. You would do this as follows:

```{r Intro 8}
mathematicians <- c("Goedel", "Euler", "Gauss", "Hilbert", "Goedel", "Fermat", "LaGrange", "Gauss")

mathematicians[stri_duplicated(mathematicians)]
```

#### Problem 8.

Suppose you have string listing famous nurses and you want to find the duplicates. Here is the list:

```{r Problem 8}
nurses <-c("Nightingale", "Barton", "Dix", "Sanger", "Barton", "Woodruff", "Lincoln", "Dix", "Peplau")
```

#### Answer 8.

```{r Answer 8}
#Insert your code to address the problem here
nurses[stri_duplicated(nurses)]
```

### Section 9.

The `LETTERS` object is a vector of length 26, consisting of all of the capital letters. Suppose we wanted to use this object to create the string`A-B_C-D_E-F_G-H_I-J_K-L_M-N_O-P_Q-R_S-T_U-V_W-X_Y-Z_`. To achieve this, we  use the following commands:

```{r Intro 9}
stri_join(LETTERS, separators = c("-","_"), collapse = "")
```

#### Problem 9.

Suppose we create the object `DIGITS` as follows:

```{r Problem 9}
DIGITS <- c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9")
```

How can we form the string `0_1-2_3-4_5-6_7-8_9-'?

#### Answer 9.

```{r Answer 9}
#Insert the code to solve the problem here.
stri_join(DIGITS, separators = c("_","-"), collapse = "")
```

### Section 10

Suppose we want to replace `statistician` with `mathematician` and `average` with `median` in the following pun:

```{r Intro 10}
pun <- "A statistician can have his head in an oven and his feet in ice, and he will say that on the average he feels fine"

pun

punModified <- 
  stri_replace_all_fixed(pun, 
                         c("statistician", "average"), 
                         c("mathematician", "median"), 
                         vectorize_all = FALSE)

punModified
```

#### Problem 10

Using the original pun, replace `his` with `her` and ` he ` with ` she `:

```{r Problem 10}
pun
```

#### Answer 10 

Note that the word `head` also starts with `he` but we don't want to substitute `she` into there and end up with `shead`.

HINT: Pay attention to spaces!!

```{r Answer 10}
#Insert the code to answer the problem here.
punModified <- 
  stri_replace_all_fixed(pun, 
                         c("his", " he "), 
                         c("her", " she "), 
                         vectorize_all = FALSE)

punModified
```

## BONUS 2: Split text into separate columns

The builtin `mtcars` dataset has row names for each car's make and model. Here is an example of the top 6 rows:

```{r}
head(mtcars)
```

Notice the row labels in the first column. Let's add these text "strings" to the dataset:

```{r}
library(dplyr)
makemodel <- row.names(mtcars)
mtcars2 <- mtcars %>%
  mutate(makemodel = makemodel)

# remove row names
row.names(mtcars2) <- NULL

# view top 6 rows again
head(mtcars2)
```

Suppose we now want to break up the make and model into separate columns using the space as our column divider. We can use the `separate()` function from `tidyr` package to do this. Note: given the full list of makes and models some have 2 spaces so you'll end up with 3 columns that we'll call "make", "model" and "type" which is why `into = c("make", "model", "type")` in the code below. This defines the new columns we are adding to the dataset.

```{r bonus2}
library(tidyr)

df <-
  tidyr::separate(
    data = mtcars2,
    col = makemodel,
    sep = " ",
    into = c("make", "model", "type"),
    remove = FALSE
  )

df %>% knitr::kable()
```

You try it. Here is a small hypothetical dataset from a lab that created custom IDs to track the subject, visit number and year by combining them into one long "string" (text field) separated by underscores "_". This is the variable `idlong` in the `labdata` dataset.

Using the code example above, see if you can (1) separate the long string `idlong` into 3 new columns added to the `labdata` dataset individually for "ID", "visit" and "year".

```{r}
# create hypothetical dataset
idlong <- c(
  "001_v1_2020",
  "001_v2_2021",
  "002_v1_2020",
  "002_v2_2021",
  "003_v1_2020",
  "003_v2_2021",
  "004_v1_2021",
  "004_v2_2022",
  "005_v1_2021",
  "005_v2_2022"
)

values <- c(34, 31, 28, 26, 34, 34, 27, 28, 30, 25)

labdata <- data.frame(idlong, values)

labdata
```

## BONUS 2 ANSWER

```{r bonus2code}
# PUT CODE FOR BONUS 2 HERE
df <-
  tidyr::separate(
    data = labdata,
    col = idlong,
    sep = "_",
    into = c("ID", "visit", "year"),
    remove = FALSE
  )

df %>% knitr::kable()
```

