---
title: "NRSG 741 - Homework 2 - Exploratory Data Analysis"
author: "YOUR NAME HERE"
date: "ADD THE DATE"
output: html_document
editor_options: 
  chunk_output_type: console
---

## INSTRUCTIONS

* Use this Rmarkdown file `N741Spring2022_Homework02_template.Rmd` to get started.
* SAVE this file under a NEW filename like `N741Spring2022_Homework02_myname.Rmd`
* Change the author to YOUR NAME
* Change the date
* Note: This Rmarkdown file has one R code chunk at the top that reads in the dataset and loads the R packages you will need.
* After each question below, insert an R code chunk to enter the R code needed to answer that question. Do this for each question.
* Outside of the R code chunk, type in any text needed to provide explanation or answer the questions further.

Note: As you Knit your document, if you are getting errors, just comment out any code that is not fully running yet by adding a `#` at the beginning of that line of code. Fix the errors and then try knitting again.

**Note: All you need to do is correctly fill in the blanks `___` in the code chunks below.**

## Homework 02 Dataset

For this Homework you will be working with the abalone dataset from the UCI (University of California - Irvine) Machine Learning data repository website: [https://archive.ics.uci.edu/ml/datasets/abalone](https://archive.ics.uci.edu/ml/datasets/abalone).

Take a look at the website to read up on the source of data and description of the variables and data in the file.

**NOTICE:** In the code below the data are actually being read directly off the internet from the UCI repository.

```{r setup, include=FALSE}
# set up chunk output options
# set to FALSE to clean up final output
# leave TRUE to help with debugging initially
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

# IGNORE THESE 2 LINES
#knitr::opts_chunk$set(error = TRUE)
#knitr::opts_chunk$set(eval = FALSE)

# Get data
# learn more at https://archive.ics.uci.edu/ml/datasets/abalone

# Load the abalone dataset using read_csv() function
# from readr package
library(readr)
abalone <- 
  readr::read_csv(file = "https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data",
                  col_names = FALSE)

# put in the variables names
# see details at https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names
names(abalone) <- c("sex","length",
                    "diameter","height",
                    "wholeWeight","shuckedWeight",
                    "visceraWeight","shellWeight","rings")

# Load dplyr and ggplot2 packages
library(dplyr)
library(ggplot2)
```

## Goal of Homework 2

This homework is meant to further your `dplyr` and `ggplot2` skills and give you a chance to do some exploratory data analysis (EDA). 

## Modify R code chunks

In each of the R code chunks below, "scaffolding" is provided. Everywhere you see 3 underscores ___ , you will need to fill in the appropriate code, variable name, function name, etc.

Use tools within the `dplyr` package as much as possible to answer the following questions.

##### **Question 1:** What kind of R object is the `abalone` dataset?

```{r q1}
# insert R code here to answer question 1
# HINT: The name of the dataset is abalone

class(abalone)
```

_TYPE ANSWER TO QUESTION 1 HERE_

ANSWER: Since this `abalone` datafile was read in using the `readr` package which is part of the tidyverse, the `abalone` dataset is a `data.frame` but more specifically it is a `"spec_tbl_df" "tbl_df"      "tbl"         "data.frame"`.

##### **Question 2:** How many observations are in the `abalone` dataset?

```{r q2}
# HINT: there are multiple ways to answer this - pick one
dim(abalone)
str(abalone)
glimpse(abalone)
nrow(abalone)
```

_TYPE ANSWER TO QUESTION 2 HERE_

ANSWER: 4177 observations

##### **Question 3:** For diameter, how many abalones have diameters less than 0.5mm?

```{r q3, eval=FALSE}
# the variable name is diameter
abalone %>%
 filter(diameter < 0.5) %>%
 nrow()
```

_TYPE ANSWER TO QUESTION 3 HERE_

ANSWER: Out of the 4177 abalones, there were 3388 with diameter measurements < 0.5mm.

##### **Question 4:** How many abalones have shucked weights larger than their whole weight? 

NOTE: There should be NO measurements where the shucked weight is > whole weight. If there are some these are probably data entry errors in this dataset.

```{r q4}
# HINT: Use a logical expression inside a filter step
# HINT: Check the spelling and case for the 
# variable names for shucked weight and whole weight
abalone %>%
 filter(shuckedWeight > wholeWeight) %>%
 nrow()
```

_TYPE ANSWER TO QUESTION 4 HERE_

ANSWER: There are 4 abalones with shucked weight higher than their whole weight, which is probably due to a data entry error since the whole weight include shucked weight. So, the whole weights should always be more than their shucked weights.

---

Create a subset containing only infants `sex == "I"`. Call this new dataset `infants`

```{r infants}
# HINT: Put the logical statement inside the filter() function
# Dont forget to use the assign operator <- to create the infants object
infants <- abalone %>%
  filter(sex == "I")
```

##### **Question 5:** How old are the abalones?

Based on the documentation provided at [https://archive.ics.uci.edu/ml/datasets/abalone](https://archive.ics.uci.edu/ml/datasets/abalone), the age of the abalones can be computed based on the number of rings where age (in years) = rings + 1.5.

Using the code scaffolding (___) fill in the blacks below to compute and add a variable called `age` to the `abalone` dataset.

Then get the summary statistics for `age`.

```{r q5}
# add age variable to abalone dataset
abalone <- abalone %>%
  mutate(age = rings + 1.5)

# select age and get summary stats
abalone %>%
  select(age) %>%
  summary()
```

What is the average age of the abalones?

_TYPE ANSWER TO QUESTION 5 HERE_

ANSWER: The average age of the abalones is 11.43 years old.

---

Show off your `dplyr` skills with `group_by()`.

##### **Question 6:** What is the average whole weight for each abalone sex (get whole weight means for females "F", males "M" and infants "I" separately)?

```{r q6}
# Hint: put the variables used in the select statement
# wholeWeight and sex.
# Remember to put
# in a name for the output of the mean() function
# something like meanwt
abalone %>%
  select(sex, wholeWeight) %>%
  group_by(sex) %>%
  summarise(meanwt = mean(wholeWeight, na.rm=TRUE))
```

_TYPE ANSWER TO QUESTION 6 HERE_

ANSWER: Table of average whole weights of abalones by sex.

```{r}
abalone %>%
  select(sex, wholeWeight) %>%
  group_by(sex) %>%
  summarise(meanwt = mean(wholeWeight, na.rm=TRUE)) %>%
  knitr::kable(caption = "Average Whole Weights by Sex")
```


##### **Question 7:** Get the means for the abalone length and height by sex?

Learn more about the `dplyr::summarise_all()` function at [https://dplyr.tidyverse.org/reference/summarise_all.html](https://dplyr.tidyverse.org/reference/summarise_all.html) - scroll down and look through the examples provided.

```{r q7}
# Hint: put variable names in the select statement
# put the function name for the mean in the 
# summarise_all() function
abalone %>%
  select(sex, length, height) %>%
  group_by(sex) %>%
  summarise_all(mean, na.rm=TRUE)
```

_TYPE ANSWER TO QUESTION 7 HERE_

ANSWER: Table of average whole weights of abalones by sex.

```{r}
abalone %>%
  select(sex, length, height) %>%
  group_by(sex) %>%
  summarise_all(mean, na.rm=TRUE) %>%
  knitr::kable(caption = "Average Lengths and Heights by Sex")
```


##### BONUS - Make a TABLE using the `arsenal` package.

Here is some example code showing the summary statistics by `sex` (M, F, I) for `length`, `diameter`, and `height` for younger abalones with `age < 12`. Use the `arsenal` package and the `tableby()` function. First create a dataset with the smaller abalones (`age < 12`) - then use that as the dataset for the `tableby()` function.

```{r results = "asis"}
library(arsenal)
abalone_young <- abalone %>%
  filter(age < 12)
  
t1 <-
  tableby(sex ~ length +
            diameter +
            height,
          data = abalone_young)

summary(t1)
```

Using the code above as a guide, now create a table showing the summary statistics by `sex` (M, F, I) for `shuckedWeight`, `visceraWeight`, and `shellWeight` for older abalones with `age >= 12`.

```{r results = "asis"}
abalone_older <- abalone %>%
  filter(age >= 12)
  
t2 <-
  tableby(sex ~ shuckedWeight +
            visceraWeight +
            shellWeight,
          data = abalone_older)

summary(t2)
```

----

## Test your graphing skills using `ggplot2`

Using the `abalone` dataset, create the following graphics/figures using `ggplot()` and associated `geom_xxx()` functions.

##### **Question 8:** Create a histogram of abalone whole weight

BONUS: Outline the histogram bars with a black line and fill the histogram bars with a green color

HINT: See example `ggplot2` code for creating histograms at:

* [http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/](http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/)
* [https://www.r-graph-gallery.com/histogram.html](https://www.r-graph-gallery.com/histogram.html)

```{r q8}
# Hint: the first option in the ggplot() function 
# is the name of the dataset
# the variable name is put inside aes()
ggplot(abalone, aes(wholeWeight)) +
  geom_histogram()

# BONUS:
ggplot(abalone, aes(wholeWeight)) +
  geom_histogram(color = "black", fill = "green")
```

_What do you notice about the distribution (any outliers or skewness)?_

_TYPE ANSWER TO QUESTION 8 HERE_

ANSWER: The distribution is right skewed. There may be a few outliers with whole weights > 2.5 or so.


##### **Question 9:** Create side-by-side boxplots of the number of rings by sex and color the bars by sex

_HINT use `geom_boxplot` with `fill=sex`_ 

See example `ggplot2` codes for boxplots at:

* [http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/](http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/) - scroll down to see boxplot examples
* [https://www.r-graph-gallery.com/boxplot.html](https://www.r-graph-gallery.com/boxplot.html)

```{r q9}
ggplot(abalone, aes(x=sex, y=rings, fill=sex)) + 
  geom_boxplot()
```

##### **Question 10:** Create a scatterplot of the whole weight on the X axis and shucked weight on the Y axis and color the points by sex

See example `ggplot2` codes for scatterplots at:

* [http://www.cookbook-r.com/Graphs/Scatterplots_(ggplot2)/](http://www.cookbook-r.com/Graphs/Scatterplots_(ggplot2)/)
* [https://www.r-graph-gallery.com/scatterplot.html](https://www.r-graph-gallery.com/scatterplot.html)

```{r q10}
ggplot(abalone, aes(x=wholeWeight, y=shuckedWeight, color=sex)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1)
```

Can you see which abalones have shucked weights > whole weights which should not happen? Look at the y=x reference line. What sex are the abalones with the incorrect weights?

_TYPE ANSWER TO QUESTION 10 HERE_

ANSWER: The 4 outliers we saw above that had shucked weights > whole weights which shouldn't happen were all for the infants which are smaller overall.

## Final Instructions

* KNIT this RMD file to PDF (or to HTML or DOC and save as PDF)
* Upload your PDF document to Canvas
